A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-21 text/introduction.tex
Shorten the introduction.

% \subsubsection{多段階計算とは何か?}

\AI{前の版は，まだ MetaOCaml specific な話と，MSP 一般の話が交互に出てきて
わかりづらかったと思います．}

@@

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with constructs for multi-stage
programming, including brackets, escape, run, and cross-stage
persistence (CSP) .  Brackets, written \verb|.< e >.| in MetaOCaml,
makes a value that represents the code (or abstract syntax tree) of
expression \verb|e|; and escape, written \verb|.~ e| and is supposed
to appear inside brackets, evaluates \verb|e| to a code value and
expands the code into the surrounding code.
For example,
\begin{verbatim}
   let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.
\end{verbatim}
evaluates to \verb|.< (fun x -> x + 1) 2 >.|.  
Run, written \verb|run e|, evaluates \verb|e| to a code value
and executes it.  For example,
\begin{verbatim}
   run (let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.)
\end{verbatim}
yields \verb|3| (here, escape connects tighter than application).
Finally, CSP is a primitive to embed values (not necessarily code
values) into a code value.  In MetaOCaml, CSP is implicitly applied to
occurrences of variables defined outside brackets.  For example,
\begin{verbatim}
   let plusone = fun x -> x + 1 in
   let y = plusone 2 in  .< y * 2 >.
\end{verbatim}
evaluates to \verb|.< 3 * 2 >.|, not \verb|.< plusone 2 * 2 >.|.
This is because the variable \verb|y| is bound to an integer \verb|3|
before \verb|.<y>.| is evaluated and the value of \verb|y| is embedded
into the code value.  In MetaOCaml, CSP can be applied to (variables of) any type.  CSP is practically important because it enables us to use library functions in brackets, e.g., \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

MetaOCaml is also equipped with a type system for safe code generation
and execution.  The notion of code types is introduced to prevent code
values that represent ill-typed expressions from being generated.  For
example,
\begin{verbatim}
   let plusone = .< fun x -> x + 1 >. in .< .~plusone "foo" >.
\end{verbatim}
is rejected by the MetaOCaml typechecker as \verb|plusone| is given
type \verb|(int -> int)| \verb|code| but it is used in the code context that
requires a function from \verb|string| is expected.  Ensuring safety
for \verb|run| is more challenging because basic code types do not
guarantee code values to be closed.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem and developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values with a type soundness theorem.

However, the type system, which is based on Hindley--Milner
polymorphism~\cite{}, is not strong enough to guarantee invariants
beyond simple types.  For example, Kiselyov~\cite{} demonstrates
specialization of vector/matrix computation with respect to the sizes
of vectors and matrices but such specialized functions can be applied
to vectors and matrices of different sizes.

@@

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specilization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes an integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector|$n$\verb| -> Vector |$n$\verb|) code|.

>>>>

@@

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with special constructs for
multi-stage programming, including brackets and espace, which are
(hygienic) quasi-quotation, and \texttt{run}, which is similar to
\texttt{eval} in Lisp, and cross-stage persistence
(CSP)~\cite{MetaML}.  Programmers can easily write code generators by
using these features.  Moreover, MetaOCaml is equipped with a powerful
type system for safe code generation and execution.  The notion of
code types is introduced to prevent code values that represent
ill-typed expressions from being generated.  For example, a quotation
of expression \texttt{1 + 1} is given type \texttt{int code}; a
quotation of \texttt{abs} is given type \texttt{(int -> int) code};
and a code-generating function, which takes a code value \(c\) as an
argument and returns \(c \texttt{ + } c\), is given type \texttt{int
  code -> int code} so that it cannot be applied to, say, a quotation
of \texttt{"Hello"}, which is given type \texttt{string
  code}.  Ensuring safety for \verb|run| is more challenging because
basic code types do not guarantee that execution of code values never
results in unbound variable errors.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem and developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values with a type soundness theorem (for a pure
subset of MetaOCaml).

% rackets, written \verb|.< e >.| in
% MetaOCaml, makes a value that represents the code (or abstract syntax
% tree) of expression \verb|e|; and escape, written \verb|.~ e| and is
% supposed to appear inside brackets, evaluates \verb|e| to a code value
% and expands the code into the surrounding code.  For example,
% \begin{verbatim}
%    let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.
% \end{verbatim}
% evaluates to \verb|.< (fun x -> x + 1) 2 >.|.  
% Run, written \verb|run e|, evaluates \verb|e| to a code value
% and executes it.  For example,
% \begin{verbatim}
%    run (let plusone = .< fun x -> x + 1 >. in .< .~plusone 2 >.)
% \end{verbatim}
% yields \verb|3| (here, escape connects tighter than application).
% Finally, CSP is a primitive to embed values (not necessarily code
% values) into a code value.  In MetaOCaml, CSP is implicitly applied to
% occurrences of variables defined outside brackets.  For example,
% \begin{verbatim}
%    let plusone = fun x -> x + 1 in
%    let y = plusone 2 in  .< y * 2 >.
% \end{verbatim}
% evaluates to \verb|.< 3 * 2 >.|, not \verb|.< plusone 2 * 2 >.|.
% This is because the variable \verb|y| is bound to an integer \verb|3|
% before \verb|.<y>.| is evaluated and the value of \verb|y| is embedded
% into the code value.  In MetaOCaml, CSP can be applied to (variables of) any type.  CSP is practically important because it enables us to use library functions in brackets, e.g., \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

However, the type system, which is based on the Hindley--Milner
polymorphism~\cite{Milner78JCSS}, is not strong enough to guarantee
invariants beyond simple types.  For example, Kiselyov~\cite{8384206}
demonstrates specialization of vector/matrix computation with respect
to the sizes of vectors and matrices in MetaOCaml but the type system
cannot prevent such specialized functions from being applied to
vectors and matrices of different sizes.

@@

One natural idea to address this problem is the introduction of dependent types
to express the size of data structure in static types~\cite{Xi98}.
% Dependent types are types which are dependent on values.
% We can use dependent types for securer programming like Xi and Pfenning
For example, we could declare vector types indexed by the size of
vectors as follows.
\begin{verbatim}
    Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor that takes an integer (which
represents the length of vectors): for example, \verb|Vector 3| is the
type for vectors whose lengths are 3.  Then, our hope is to specialize
vector/matrix functions with respect to their size and get a piece of
function code, whose type respects the given size, \emph{provided at
  specilization time}.  For example, we would like to specialize a
function to add two vectors with respect to the size of vectors, that
is, to implement a code generator that takes an integer $n$ as an
input and generates a piece of function code of type
\verb|(Vector |$n$\verb| -> Vector |$n$\verb| -> Vector |$n$\verb|) code|.
